Gradual Soundness: Lessons from Static Python

### 2022-01-28 pj-7.1/paper.tex
pj: 22 pages finally no thanks to baby cub

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\PEP}{PEP~484}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{~\mid~} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}
\newcommand{\nnum}[1]{$N=#1$}
\newcommand{\totalnum}[1]{$#1$ total}
\newcommand{\pctnum}[1]{$#1\%$}

@@

\subsection{Type System Context and Design Goals}

@@

Cinder is a bytecode compiler that extends the Python instruction set.
All the standard Python instructions work similarly to Python 3.8;
in other words, an untyped Python program has the same behavior in
Cinder as in the standard interpreter.
The added instructions help with one of three tasks:
expressing runtime checks,
initializing concrete-typed values,
or efficiently performing a standard action.
As an example of the third kind, the \code{FAST\_LEN} instruction quickly
computes the length of a built-in value.
\SP{} uses type information to decide where this instruction is appropriate
and applies it as an optimization.

@@

Cinder adds virtual method tables (vtables) to classes.
These tables help to speed up method dispatch relative to Python's
dynamic lookup.
Calls to static methods that appear in statically-typed
code use the vtable to find an address for the method.
If the method is part of a final class, then the call is further optimized
to a direct function-call jump.
(Both vtable lookup and direct jumps are supported by Cinder-specific bytecode
instructions.)
Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
function calls.

The implementation of vtables happens to be built on the Python \code{vectorcall}
protocol; it is not a from-scratch development.
Cinder vtables are specialized to check untyped overrides of typed methods
using a wrapper implemented in C rather than in Python, to reduce performance costs.

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All untyped writes must be protected by casts.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a compound type tag 
    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
    to support casts from the dynamic type.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{K'}{V'}$ only if $K$ is equal to $K'$
    and $V$ is equal to $V'$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

@@

Cinder includes a JIT (just-in-time) compiler for its bytecode.
At the moment, programmers enable the JIT by supplying a list of method names
to the compiler.
Any method can be JIT-compiled whether or not it uses \SP{} types,
though types are needed to enable the most-effective optimizations.

@@

To give one example issue, \SP{} accepted the following program whereas our model
reported a type error:
%% \url{https://github.com/facebookincubator/cinder/issues/62}

@@

\noindent{}The function \code{f} expects either a string or the \code{None} value and promises to return a string.
When called with \code{None}, however, the function breaks out of the \code{while} loop
and implicitly returns \code{None} contrary to the return type.
\SP{} had failed to account for the \code{break} and implicit return.
More concerningly, the associated test case \emph{expected} the program to type check.
Creating a formal model helped us catch this specification error.

@@

\Cref{f:surface-types} presents a core surface syntax for model programs.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, integers,
booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.

In Python, the syntax for expressions overlaps with the syntax of types.
The Python name \code{None} is both an expression and a type,
and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
names (\code{None}, \code{CheckedDict}).

@@

\SP{} evaluation types can all be enforced fully and efficiently
with casts.
\Cref{t:cast} describes the casts in detail.
Most call for a tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Even checked dictionaries rely on tag checks.
A checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, the exact-match rules applies even when $\spteval_0$
or $\spteval_1$ is the dynamic type.

No cast requires traversing a data structure.
Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
This latter property means that blame-tracking~\cite{ff-icfp-2002} is trivial because
casts either succeed or fail immediately.
%% In other words, \SP{} provides \emph{immediate
%% accountability}~\cite{mt-oopsla-2017} for cast errors.

@@

To better show where \SP{} needs to insert casts, we present a selection of typing rules.
Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
These declarations fill a type environment ($\sptenv$).
Relative to the current environment, the expressions within each statement must
satisfy the typing judgment in~\cref{f:eval-types}.

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~(\cref{s:c-types}), module-level variables,
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.

@@

%% TODO FILL only 2 cases? revisit after adding "changes to python" section in 2
There are two cases in which \SP{} types impose constraints on untyped Python code.
First, concrete generic types (e.g., $\sptrawchkdict$)
reject inputs that were not created via a checked constructor.
If a client of a Python module decides to impose concrete types, the Python module
may need to change.
Second, typed modules prevent updates to module-level variables.

Interactions with {open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs;
the optimizer can bypass these checks for typed-to-typed calls~(\cref{s:optimize}).
Similarly, concrete types such as $\sptrawchkdict$ must check writes that
come from untyped contexts.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness

@@

\SP{} can generate efficient code because it targets
the Cinder runtime.
Cinder extends the Python 3.8 bytecode with instructions to check types at run-time,
to construct \SP{} data structures, and to perform actions in a way that benefits from
static information.

@@

\Cref{t:bytecode} lists a few representative bytecode instructions.
The two instructions that express run-time checks are \code{CAST}
and \code{CHECK\_ARGS}.
The former checks a value against a type.
The latter is for functions and methods; it checks all inputs to a function
against their declared types.

Two of the listed instructions allocate \SP{}-specific data structures.
The first creates a checked dictionary; the second allocates an uninitialized instance
of an object~(\cref{s:vtable}).
Cinder comes with a similar instruction to build checked lists.
%% The developers will add more instructions along the same lines in the future.

The remaining instructions are for optimization.
Both \bcinst{INVOKE\_FUNCTION} and \bcinst{INVOKE\_METHOD} are alternatives
to Python's dynamic call dispatch.
The former uses the classloader to quickly find the address of a function;
the latter uses a vtable lookup to find a method.
The load and store instructions for fields improve upon Python's generic attribute lookup.
In the JIT, these instructions can sometimes be further optimized to a single assembly instruction.
The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
of using the Python \code{\_\_len\_\_} attribute.
Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value whose
type is not clear from the context.

\paragraph{Removing Casts}

@@

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialize rule in the model illustrates this policy, which ensures that
well-typed writes to classes and to concrete dictionaries have no cost.
Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
that defensively casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; namely, whenever a typed function or method gets
called in a typed context.

@@

To measure the effect of a code change on these hosts, profiling tools select
two representative sets of machines to run as experimental and control groups.
The experimental groups gets the latest version of the server code;
the control group gets the previous version.
Next, the profiling tools slowly increase the number of requests that these
machines receive until each is running at maximum load.
Once the machines are fully allocated, the tools measure how many requests
per second each group is able to handle.
If the experimental group can serve more requests per second, then the change
is recorded as a success.
Assuming that the experimental and control groups have equivalent hardware
capabilities, and assuming that both groups receive a representative sample
of requests, then the results of an experiment (increase/decrease in CPU efficiency)
should predict actual performance.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython improved performance by compiling these modules to C,
but its partial adoption had led to an awkward workflow (with a
few compiled modules spread across an interpreted codebase).
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

@@

In general, the patches have minor code changes and most of these changes
affect tests rather than production code.
Digging deeper, there are nine common kinds of changes: six address
type errors and the other three are for performance.

@@

\begin{enumerate}
  \item
    \ipara{Use primitive types}
    Any Python integers and booleans in hot code paths are better as
    \SP{} primitive values~(\cref{s:c-types}).
    As noted above, primitives were required for \SP{} to match the performance of
    Cython.
    Twelve modules currently use primitives.
    %% - bg: focus on block, convert, no problem?

@@

  \item
    \ipara{Change functions to accept positional arguments}
    Functions calls that use only keyword arguments are not yet optimized by \SP{}.
    Rewriting these functions and call-sites to use positional arguments improves performance.

@@

\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.\footnote{\url{https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks}}
These microbenchmarks are admittedly small and quite different than typical application code,
but they provide an auditable and reproducible way to measure performance.

@@

All runtimes for this table came from a 64-bit CentOS Linux system using a build
of \SP{} and Cinder compiled with default settings
(commit \href{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}{\code{d0d071a}}).
Each benchmark version ran 11 times total: once to warm up the bytecode caches
(\code{.pyc} files) and 10 times to compute an average runtime (using the
\code{time} utility).
Every benchmark version invoked the JIT on the same functions---even the
untyped versions.

@@

Almost all ratios are less than one, showing that \SP{} achieves a speedup over
Python despite the cost of enforcing soundness.
The one exception is the basic typed version of \bmname{fannkuch}, which sees a minor
slowdown.
This benchmark is primarily about number-crunching and the basic version uses standard
numeric types that \SP{} does not optimize.
Switching to primitive types without otherwise changing the code
gives an immediate speedup (ratio = $0.90$).
\Cref{a:microbench-detail} presents further benchmark details.

@@

Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wnlov-popl-2010}
and StrongScript~\cite{rzv-ecoop-2015},
and play a central role in Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021}.
It was thanks to Nom that \SP{} acquired concrete types; the impressive performance results
in the original Nom paper inspired the \SP{} team to pursue an implementation.

In addition to concrete types, Thorn and StrongScript have optional like types.
The mix of like and concrete types is similar to \SP{}'s mix of concrete and shallow
types, though like types make no claims about soundness.

Many aspects of our Redex model for \SP{} are adapted from the Full Monty core calculus
for Python~\cite{pmmwplck-oopsla-2013}.
In particular, the Full Monty paper reminded us that booleans \emph{are} integers in Python;
\SP{} had overlooked this detail.\footnote{Bug report: \url{https://github.com/facebookincubator/cinder/issues/46}}
%% 2022-01-14 TODO check full monty with KC

Our production experience with \SP{} suggests that a method-based JIT compiler can
usually eliminate the costs of shallow types.
This finding complements prior positive experience with shallow types and JIT
technology; e.g., in Reticulated with the pypy tracing JIT~\cite{vsc-dls-2019}
and in Grace with the Graal partial-evaluation-based JIT~\cite{rmhn-ecoop-2019,grmhn-vmil-2019}.
Similarly, the Pycket team used a tracing JIT to reduce the costs of higher-order
checks~\cite{bbst-oopsla-2017}.
It remains to be seen whether a non-tracing JIT is effective for higher-order
casts.

\SP{} is one of many type systems for Python.
Other optional type checkers include Pyre~\cite{pyre},
mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}; these
systems do not ensure safe interactions with untyped Python.
Another sound type system is Reticulated,
which enforces structural types (rather than nominal) using its transient
semantics~\cite{vss-popl-2017,v-thesis-2019}.
Other \emph{optimizing} type systems for Python
include Reticulated and mypyc.
The latter compiles Python source to C extension modules; unlike \SP{},
it does not offer a pure-Python developer experience.

Although sound gradual type systems are rare in industry, there are at
least three others besides \SP{}.
Dart 2 is a nominally-typed language with a dynamic type that is similar
to the top object type but allows implicit downcasts to every other type~\cite{dart-types}.
C\# includes a dynamic type that defers type checking until runtime~\cite{bmt-ecoop-2010}.
JS++ is a sound type system for JavaScript that allows untyped values
using a catch-all \code{external} type~\cite{jspp}.
%% Programmers can manually cast external values to a more precise type.
%% The non-immediate casts make copies (how does that work for functions?!)
%% also, how does JS++ protect arrays sent to JS?

Gradual soundness in \SP{} bears a strong resemblance to \emph{progressive types} vision
of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
Primitive types are progressive types for base values.
Concrete types are essentially a progressive form of shallow types.

@@

\SP{} is an ambitious undertaking.
Its developers are maintaining a fork of the Python 3.8 runtime,
a type checker,
an optimizing ahead-of-time compiler,
and a method-based JIT compiler.
The team has converted hundreds of untyped modules to \SP{}
and plans to convert thousands more in the future.

And yet, \SP{} is carefully restricted to set attainable goals
for its implementors and for adopters:
\begin{itemize}
  \item
    The syntax of sound static types is relatively small compared
    to the \PEP{} definition.
    Higher-order types are absent.
    The focus is instead on first-order class types, base types, and standard data structures.
  \item
    The type checker operates at a module-level granularity rather
    than at the level of statements or expressions.
  \item
    Gradual class hierarchies are permitted, but restricted to single inheritance.
    Furthermore, static-to-static method overrides cannot reduce the precision of types---a restriction
    that enables optimizations but breaks the static gradual guarantee.
  \item
    Shallow types such as $\sptrawpydict$ accommodate Python code but offer coarse
    soundness guarantees.
    These types have been the top priority for \SP{}.
    Concrete types support full soundness, but impose nonlocal maintenance costs.
    \SP{} supports few concrete types at the moment and plans to add more.
  \item
    Progressive primitive types are available to help developers maximize performance.
    These types are intentionally incompatible with the dynamic type to help programmers
    avoid casts.
\end{itemize}
%
Many of these restrictions go against the grain of the mainstream
research community, but they have been effective in practice.
The current \CPUchange{} increase in CPU efficiency is a huge improvement
at the scale of the Instagram web server.

In conclusion, we conjecture that the \SP{} approach to gradual typing
is a viable way to incrementally add types to a production codebase.
Determining whether this conjecture holds more broadly is a critical
question that calls for research on two fronts.
First, \SP{} must be applied to additional projects in a module-at-a-time manner.
A exhaustive performance evaluation~\cite{gtnffvf-jfp-2019} of a few small projects would be
especially useful.
Second, the \SP{} approach should be adapted to other languages.
An optionally-typed language such as TypeScript would be an ideal starting
point for such an experiment.
Gradual, progressive soundness may prove itself to be an effective general strategy for
gradual types in practice.

>>>>

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\JSPP}{JS\code{++}}
\newcommand{\PEP}{PEP~484}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{~\mid~} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}
\newcommand{\nnum}[1]{$N=#1$}
\newcommand{\totalnum}[1]{$#1$ total}
\newcommand{\pctnum}[1]{$#1\%$}

@@

\subsection{Type System Context and Design Goals}
\label{s:ts-context}

@@

All CPython 3.8 instructions have the same behavior in Cinder.
Cinder adds instructions to help with one of three tasks:
expressing runtime checks,
initializing concrete-typed values,
or efficiently performing a standard action.
As an example of the third kind, the \code{FAST\_LEN} instruction quickly
computes the length of a built-in value, which helps to optimize for loops.
\SP{} uses type information to decide where this instruction is appropriate
and applies it as an optimization.

@@

Cinder adds virtual method tables (vtables) to classes.
These tables help to speed up method dispatch relative to Python's
dynamic lookup.
Calls to static methods that appear in statically-typed
code use the vtable to find an address for the method.
If the method is part of a final class, then the call is further optimized
to a direct function-call jump.
%% Just-in-time compilation~(\cref{s:jit}) can also upgrade vtable lookups to
%% function calls.
(Both vtable lookup and direct jumps are supported by Cinder-specific bytecode
instructions.)

The implementation of vtables happens to be built on the Python \code{vectorcall}
protocol; it is not a from-scratch development.
Cinder vtables are further specialized to check untyped overrides of typed methods
using a wrapper implemented in C rather than in Python, to reduce performance costs.

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All untyped writes must be protected by casts.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a compound type tag 
    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
    to support casts from the dynamic type.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{K'}{V'}$ only if $K$ is equal to $K'$
    and $V$ is equal to $V'$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a compound tag.

@@

Cinder includes a JIT (just-in-time) compiler for its bytecode.
At the moment, programmers enable the JIT by supplying a list of method names
to the compiler.
Any method can be JIT-compiled whether or not it uses \SP{} types,
although types are needed to enable the most-effective optimizations.

@@

To give one example issue, \SP{} incorrectly accepted the following program whereas our model
reported a type error:
%% \url{https://github.com/facebookincubator/cinder/issues/62}

@@

\noindent{}The function \code{f} expects either a string or the \code{None} value and promises to return a string.
When called with \code{None}, however, the function breaks out of the \code{while} loop
and implicitly returns \code{None} contrary to the return type.
\SP{} had failed to account for the \code{break} and implicit return.
More concerningly, the associated test case \emph{expected} the program to type check.
Our model caught this specification error.

@@

\Cref{f:surface-types} presents a core syntax for programs.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values are the none value, integers,
booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
Other expressions describe function calls, dictionary reads and writes, object field reads and writes,
and method calls.

In Python, the syntax for expressions overlaps with the syntax of types.
For example, the Python name \code{None} is both an expression and a type,
and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
\Cref{f:surface-types} does not follow these standards; instead it keeps the two syntaxes distinct.
Expressions use lowercase names (\code{none}, \code{chkdict}) and types uses capitalized
names (\code{None}, \code{CheckedDict}).

@@

\SP{} evaluation types can all be enforced fully and efficiently
with casts.
\Cref{t:cast} describes the casts in detail.
Most call for a tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Even checked dictionaries rely on tag checks.
A checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, this exact-match rules applies even when $\spteval_0$
or $\spteval_1$ is the dynamic type.

No cast requires traversing a data structure.
Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
This latter property means that blame-tracking~\cite{ff-icfp-2002} is trivial because
casts either succeed or fail immediately.
In other words, \SP{} provides \emph{immediate accountability} same as
Nom~\cite{mt-oopsla-2017}.

@@

To show where \SP{} needs to insert casts, we present a selection of typing rules.
Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
These declarations fill a type environment ($\sptenv$).
Relative to the current environment, the expressions within each statement must
satisfy the typing judgment in~\cref{f:eval-types}.

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~(\cref{s:c-types}), module-level variables,
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components; namely, a model of the \SP{} bytecode and a faithful rendering of its
transformations.

@@

%% TODO FILL only 2 cases? revisit after adding "changes to python" section in 2
There are two cases in which \SP{} types impose dynamic constraints on untyped Python code.
First, concrete generic types (e.g., $\sptrawchkdict$)
reject inputs that were not created via a checked constructor.
If a client of a Python module decides to impose concrete types, the Python module
may need to change.
Second, typed modules prevent updates to module-level variables.

Interactions with {open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
The optimizer can bypass these checks for typed-to-typed calls~(\cref{s:optimize}).
Similarly, concrete types such as $\sptrawchkdict$ must check writes that
come from untyped contexts.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness

@@

@@

\SP{} can generate efficient code because it targets
the Cinder runtime, which has instructions to check types at run-time, to
construct \SP{} data structures, and to perform actions in a way that benefits
from static information.
\Cref{t:bytecode} lists a few representative bytecode instructions.
The two instructions that express run-time checks are \bcinst{CAST}
and \bcinst{CHECK\_ARGS}.
The former checks a value against a type.
The latter is for functions and methods; it checks all inputs to a function
against their declared types.

The instructions \bcinst{BUILD\_CHECKED\_MAP} and \bcinst{TP\_ALLOC}
allocate \SP{}-specific data structures.
The first creates a checked dictionary; the second allocates an uninitialized instance
of an object~(\cref{s:vtable}).
Cinder comes with a similar instruction to build checked lists.
%% The developers will add more instructions along the same lines in the future.

The remaining instructions are for optimization.
Both \bcinst{INVOKE\_FUNCTION} and \bcinst{INVOKE\_METHOD} are alternatives
to Python's dynamic call dispatch.
The former uses the classloader to quickly find the address of a function;
the latter uses a vtable lookup to find a method.
The load and store instructions for fields improve upon Python's generic attribute lookup.
In the JIT, these instructions can sometimes be further optimized to a single assembly instruction.
%% The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
%% of using the Python \code{\_\_len\_\_} attribute.
Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value whose
type is not clear from the context.

\paragraph{Reducing Casts}

@@

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialize rule in the model illustrates this policy, which ensures that
well-typed writes to classes and to concrete dictionaries have no cost.
Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
that defensively casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; that is, whenever a typed function or method gets
called in a typed context.

@@

To measure the effect of a code change on these hosts, profiling tools select
two representative sets of machines to run as experimental and control groups.
The experimental groups gets the latest version of the server code;
the control group gets the previous version.
Next, the profiling tools slowly increase the number of requests that these
machines receive until each is running at maximum load.
Once the machines are fully allocated, the tools measure how many requests
per second each group is able to handle.
If the experimental group can serve more requests per second, then the change
is a success.
Assuming that the experimental and control groups have equivalent hardware
capabilities, and assuming that both groups receive a representative sample
of requests, then the results of an experiment (increase/decrease in CPU efficiency)
should predict actual performance.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython improved performance by compiling these modules to C,
but its partial adoption had led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

@@

In general, the patches have minor code changes and most of these changes
affect tests rather than production code.
Digging deeper, there are nine common kinds of changes: five address
type errors and the other four are for performance.

@@

\begin{enumerate}
  \item
    \ipara{Use primitive types}
    Integers and booleans in hot code paths run fastest with
    primitive type~(\cref{s:c-types}).
    %%As noted above, primitives were required for \SP{} to match the performance of
    %%Cython.
    Twelve modules currently use primitives.
    %% - bg: focus on block, convert, no problem?

@@

  \item
    \ipara{Change functions to accept positional arguments}
    Functions calls that use only keyword arguments are not yet optimized by \SP{}.
    Rewriting these functions and their call-sites to use positional arguments improves performance.

@@

\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.
These microbenchmarks are admittedly small and quite different than typical application code,
but they provide an auditable and reproducible way to measure performance.

@@

All runtimes for this table came from a 64-bit CentOS Linux system using a build
of \SP{} and Cinder compiled with default settings.
Each benchmark version ran 11 times total: once to warm up the bytecode caches
(\code{.pyc} files) and 10 times to compute an average runtime (using the
\code{time} utility).
Every benchmark version invoked the JIT on the same set of functions.
\Cref{a:microbench-detail} presents further details.

@@

Almost all ratios are less than one, showing that \SP{} achieves a speedup over
Python despite the cost of enforcing soundness.
The one exception is the basic typed version of \bmname{fannkuch}, which sees a minor
slowdown.
This benchmark is primarily about number-crunching and the basic version uses standard
numeric types that \SP{} does not optimize.
Switching to primitive types without otherwise changing the code
gives an immediate speedup (ratio = $0.90$).

%% FILL threats to validity

@@

Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wnlov-popl-2010}
and StrongScript~\cite{rzv-ecoop-2015},
and play a central role in Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021}.
%% It was thanks to Nom that \SP{} acquired concrete types; the impressive performance results
%% in the original Nom paper inspired the \SP{} team to pursue an implementation.
In addition to concrete types, Thorn and StrongScript have optional (unsound) \emph{like} types.
The mix of like and concrete types is similar to \SP{}'s mix of concrete and shallow
types.

Many aspects of our Redex model for \SP{} are adapted from the Full Monty core calculus
for Python~\cite{pmmwplck-oopsla-2013}.
In particular, the Full Monty paper reminded us that booleans are integers in Python;
\SP{} had overlooked this detail.
%% \footnote{Bug report: \url{https://github.com/facebookincubator/cinder/issues/46}}
%% 2022-01-14 TODO check full monty with KC

Production experience with \SP{} suggests that a method-based JIT can
eliminate the costs of shallow types.
This finding complements prior work on shallow types;
e.g., in Reticulated with the pypy tracing JIT~\cite{vsc-dls-2019}
and in Grace with the Graal partial-evaluation-based JIT~\cite{rmhn-ecoop-2019,grmhn-vmil-2019}.
Similarly, the Pycket team successfully used a tracing JIT to reduce the costs of higher-order
checks~\cite{bbst-oopsla-2017}.
%% It remains to be seen whether a non-tracing JIT is effective for higher-order
%% casts.

\SP{} is one of many type systems for Python.
Other optional type checkers include Pyre~\cite{pyre},
mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}.
Another sound type system is Reticulated, which enforces structural types and
pioneered the transient semantics~\cite{vss-popl-2017,v-thesis-2019}.
Optimizing type systems include Reticulated and mypyc~\cite{mypyc}.
%% The latter compiles Python source to C extension modules; unlike \SP{},
%% it does not offer a pure-Python developer experience.

Although sound gradual types are rare in industry, there are at
least three other languages that provide them.
Dart 2 is a nominally-typed language with a dynamic type that is similar
to a top (object) type but statically allows all method calls~\cite{dart-types}.
C\# has a dynamic type that delays type checking until runtime~\cite{bmt-ecoop-2010}.
\JSPP{} is typed JavaScript that allows untyped values
using a catch-all \code{external} type~\cite{jspp}.
%% Programmers can manually cast external values to a more precise type.
%% The non-immediate casts make copies (how does that work for functions?!)
%% also, how does JS++ protect arrays sent to JS?

%% Gradual soundness in \SP{} bears a strong resemblance to \emph{progressive types} vision
%% of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
%% Primitive types are progressive types for base values.
%% Concrete types are essentially a progressive form of shallow types.

@@

\SP{} is an ambitious undertaking.
Its developers are maintaining a fork of the Python 3.8 runtime,
a type checker,
an optimizing ahead-of-time compiler,
and a method-based JIT compiler.
The team has converted hundreds of untyped modules to \SP{}
and plans to convert thousands more in the future.
And yet, \SP{} is carefully restricted to set attainable goals:
%% for its implementors and for adopters:
\begin{itemize}
  \item
    The syntax of sound static types is relatively small compared
    to \PEP{}.
    Higher-order types are absent.
    The focus is instead on first-order objects and standard data structures~(\cref{s:ts-context}).
  \item
    The type checker analyzes entire modules rather than individual statements,
    and although it supports a dynamic type, dynamic type-checked code
    does not have the same freedom as untyped code~(\cref{s:type-dynamic}).
  \item
    Shallow types accommodate Python code but offer coarse
    soundness guarantees.
    %% These types have been the top priority for \SP{}.
    Concrete types are deeply sound, but add nonlocal constraints~(\cref{s:checked-type}).
    %% \SP{} supports few concrete types at the moment and plans to add more.
  \item
    Gradual class hierarchies are permitted, but restricted to single inheritance.
    Static-to-static method overrides cannot reduce the precision of types~(\cref{s:inheritance}).
    %% ---a restriction
    %% that enables optimizations but breaks the static gradual guarantee.
  \item
    Progressive primitive types are available to maximize performance~(\cref{{s:c-types}}).
    %%These types are intentionally incompatible with the dynamic type to help programmers
    %%avoid casts.
\end{itemize}
%
Many of these restrictions go against the grain of the mainstream
research community, but they have been effective in practice.
The current \CPUchange{} increase in Instagram web server CPU efficiency is a huge improvement
and was obtained incrementally without any careful planning to avoid performance bottlenecks.
By contrast, bottlenecks are a major concern for expressive, structurally-typed
gradual languages~\cite{gtnffvf-jfp-2019,tfgnvf-popl-2016}.

In conclusion, the \SP{} approach to gradual typing seems to be a promising
way to realize the vision of gradual typing.
Determining whether this conjecture holds more broadly
calls for research on two fronts.
First, \SP{} must be applied to additional projects in a module-at-a-time manner.
A careful evaluation~\cite{gtnffvf-jfp-2019} of a few small projects would be
especially useful to find performance bottlenecks.
Second, the \SP{} approach should be adapted to other languages.
An optionally language such as TypeScript would be an ideal starting
point.
%% Gradual, progressive soundness may prove itself to be an effective general strategy for
%% gradual types in practice.
